\section{Chapter V}
\translation{
  Quae cum ita sint, Catilina, perge, quo coepisti, egredere aliquando ex urbe; patent portae; proficiscere. Nimium diu te imperatorem tua illa Manliana castra desiderant. Educ tecum etiam omnes tuos, si minus, quam plurimos; purga urbem. Magno me metu liberabis, dum modo inter me atque te murus intersit. Nobiscum versari iam diutius non potes; non feram, non patiar, non sinam. 

  11. Magna dis inmortalibus habenda est atque huic ipsi Iovi Statori, antiquissimo custodi huius urbis, gratia, quod hanc tam taetram, tam horribilem tamque infestam rei publicae pestem totiens iam effugimus. Non est saepius in uno homine summa salus periclitanda rei publicae. Quamdiu mihi consuli designato, Catilina, insidiatus es, non publico me praesidio, sed privata diligentia defendi. Cum proximis comitiis consularibus me consulem in campo et competitores tuos interficere voluisti, compressi conatus tuos nefarios amicorum praesidio et copiis nullo tumultu publice concitato; denique, quotienscumque me petisti, per me tibi obstiti, quamquam videbam perniciem meam cum magna calamitate rei publicae esse coniunctam.

  12. Nunc iam aperte rem publicam universam petis, templa deorum inmortalium, tecta urbis, vitam omnium civium, Italiam [denique] totam ad exitium et vastitatem vocas. Quare, quoniam id, quod est primum, et quod huius imperii disciplinaeque maiorum proprium est, facere nondum audeo, faciam id, quod est ad severitatem lenius et ad communem salutem utilius. Nam si te interfici iussero, residebit in re publica reliqua coniuratorum manus; sin tu, quod te iam dudum hortor, exieris, exhaurietur ex urbe tuorum comitum magna et perniciosa sentina rei publicae. 

13. Quid est, Catilina? num dubitas id me imperante facere, quod iam tua sponte faciebas? Exire ex urbe iubet consul hostem. Interrogas me, num in exilium; non iubeo, sed, si me consulis, suadeo.}{
  Since these things are thus, Catiline, proceed to where you have begun, now at last go out from the  city; the gates are open: set out. That Manlian camp of yours has been longing for you, its general,  for an excessively long time. Also, lead out with you all your men, if less, as many as possible; cleanse  the city. You will free me from a great fear, only provided that there is a wall between me and you.  You cannot dwell with us now any longer: I will not bear it, I will not endure it, I will not allow it.

  Great thanks must be given to the immortal gods and to this Jupiter Stator himself, the mostancient guardian of this city, because we have already so often escaped this scourge, so foul, sodreadful and so hostile to the republic. The utmost safety of the republic must not be so often putat risk by one man. Catiline, all the time you plotted against me, (when I was) the consul elect, Idefended myself not with a public garrison but with private vigilance. When at the most recentconsular elections you wished to kill me, the consul, and your fellow candidates on the plain (ofMars), I supressed your wicked attempts with a garrison of friends and with my troops, with no riotpublicly instigated; in short, as often as you attacked me, I prevented you through my own means,although I saw that my destruction was linked with great damage to the republic.

  Now you are already openly attacking the entire republic; you are calling for the destruction anddevastation of the temples of the immortal gods, the buildings of the city, the lives of all the citizensand the whole of Italy. Therefore, since I do not yet dare to do that which is most important andwhich is in keeping with this supreme authority and the teaching of our ancestors, I shall do thatwhich is more lenient regarding severity and more useful regarding the common safety. For if I orderyou to be killed, there will remain behind in the republic the rest of the band of conspirators; but ifyou go out, which I have been urging you (to do) for a long time now, the great dregs of yourcompanions, dangerous to the republic, will be drained out of the city.

  What is it, Catiline? Surely you are not hesitating to do at my order what you were already doingof your own accord? The consul orders the enemy to go out of the city. You put the question to me:surely not into exile? I do not order it, but if you ask my advice, I advise it.
}
