\section{Chapter VI}
\translation{Quid est enim, Catilina, quod te iam in hac urbe delectare possit? in qua nemo est extra istam coniurationem perditorum hominum, qui te non metuat, nemo, qui non oderit. Quae nota domesticae turpitudinis non inusta vitae tuae est? quod privatarum rerum dedecus non haeret in fama? quae lubido ab oculis, quod facinus a manibus umquam tuis, quod flagitium a toto corpore afuit? cui tu adulescentulo, quem corruptelarum inlecebris inretisses, non aut ad audaciam ferrum aut ad lubidinem facem praetulisti?

  14. Quid vero? nuper cum morte superioris uxoris novis nuptiis domum vacuefecisses, nonne etiam alio incredibili scelere hoc scelus cumulasti? quod ego praetermitto et facile patior sileri, ne in hac civitate tanti facinoris inmanitas aut extitisse aut non vindicata esse videatur. Praetermitto ruinas fortunarum tuarum, quas omnis inpendere tibi proxumis Idibus senties; ad illa venio, quae non ad privatam ignominiam vitiorum tuorum, non ad domesticam tuam difficultatem ac turpitudinem sed ad summam rem publicam atque ad omnium nostrum vitam salutemque pertinent.

  15. Potestne tibi haec lux, Catilina, aut huius caeli spiritus esse iucundus, cum scias esse horum neminem, qui nesciat te pridie Kalendas Ianuarias Lepido et Tullo consulibus stetisse in comitio cum telo, manum consulum et principum civitatis interficiendorum causa paravisse, sceleri ac furori tuo non mentem aliquam aut timorem tuum sed fortunam populi Romani obstitisse? Ac iam illa omitto---neque enim sunt aut obscura aut non multa commissa postea---: quotiens tu me designatum, quotiens consulem interficere conatus es! quot ego tuas petitiones ita coniectas, ut vitari posse non viderentur, parva quadam declinatione et, ut aiunt, corpore effugi! nihil agis, nihil adsequeris, neque tamen conari ac velle desistis.

  16. Quotiens tibi iam extorta est ista sica de manibus, quotiens excidit casu aliquo et elapsa est! quae quidem quibus abs te initiata sacris ac devota sit, nescio, quod eam necesse putas esse in consulis corpore defigere.
}{
  For what is there, Catiline, which can now please you in this city? There is no one in it outside that conspiracy of worthless people who doesn’t fear you, no one who doesn’t hate you. What work of domestic turpitude has not been branded onto your life? What shame of private affairs does not cling to your reputation? What lust was ever absent from your eyes, what crime was ever absent from your hands, what wickedness was ever absent from your whole body? To which mere lad did you not carry in front the sword for their recklessness or a torch for lust, whom you have ensnared with the lure of corruption?

  What indeed? Lately, when you had emptied your home from the death of your former wife for your new weddings, surely you also increased this crime with another incredible act of wickedness? Because I make no mention of it and allow it to be easily silenced, lest the monstrosity of such a great crime in this state seems to either have existed or have not been avenged. I pass over the collapse of your fortune, all of which you will realise are hanging over you at the next Ides; I come to those matters, not to the personal disgrace of your vices, nor your domestic difficulties and turpitude, but to concern the utmost danger to the republic and life and safety of us all.

  Can this light or air of this sky be pleasing to you, Catiline, when you know there to be no one of these men, who are unaware that you stood with a weapon in the Comitium the day before the Kalends of January during the consulship of Lepidus and Tullus? That you prepared your band for the sake of killing the consuls and leading citizens of the state? That not some change of mind or fear on your part stood in the way of crime and your frenzy, but the good luck of the people of Rome? And now I leave out those things (for your offences after that are neither unknown nor few): how often you tried to kill me when I was consul designate, when I was actually consul! How many of your thrusts, having been hurled in such a way, that they seem to be unable to be avoided, I evaded by a mere swerve and turn of the body, as they say! You do nothing, you achieve nothing, however you do not cease to try and wish. 

How often has that dagger been wrested from your hands! How many times has it accidentally fallen and slipped somewhere! I do not know indeed to what sacred things that dagger has been initiated and offered by you, because you think that it is necessary to thrust in the body of a consul.
}
% Now indeed what sort of life is that life of yours? For in this way now I will speak
% with you, not in order that I may seem to have been impelled by hatred, by which I
% ought to be, but by pity, nothing of which is due to you. You came a little while
% before into the senate. Who from this so great an assembly greeted you? If this after
% the memory of men happened to no one else, are you waiting for an insult with
% words, when you may be oppressed by a very heavy judgment of silence? What,
% because at your arrival these seats were vacant, because all consuls, who have been
% quite often set for slaughter by you, and at the same time you sat down, left that part
% of the benches bare and vacant, at last with which mind do you think you should bear
% this?
% }
